### 2023-07-16 tex/intro.tex
[ writing ] address all remaining todos

FIXME: the start of this chapter is directly from the ESOP22 paper.

In this paper, we treat the metatheory of a class of substructural type
systems related to linear logic~\cite{girard87linear}.
This class is variously known as
\emph{coeffectful}~\cite{PetricekOM14,Granule18},
\emph{quantitative}~\cite{BrunelGMZ14,Atkey18}, or
\emph{resource-aware}~\cite{GhicaS14},
or is given no particular name~\cite{reed10distance,abadi99core},
and generalises bounded linear logic to track variable usage with semiring
annotations.
In all of these systems, we have some ambient semiring $\Ann$, and in the
judgements of the type system, variables are annotated by elements of $\Ann$
describing \emph{how} that variable can be used.
The additive structure of $\Ann$ gives the ability to count, or otherwise
accumulate, usages of variables in multiple subterms.
The multiplicative structure gives rise to a form of modality, for example
allowing multiple or unlimited reuse, or movement between security levels, in
the type system.

The aspect of such systems we tackle here is their basic metatheory and
mechanisation thereof.

We build upon both the general structural framework of
\citet{AACMM21} and the substructural techniques of \citet{WA21}.
The way \citeauthor{AACMM21} consolidate and codify mechanisation techniques for
propositional natural deduction systems based on intrinsically typed syntax and
de Bruijn indices, we aim to replicate for linear-like systems based on
semiring usage annotations.
By picking a trivial semiring, our work can subsume that of
\citeauthor{AACMM21}, except for the many pieces of machinery we have not yet
ported to this new framework.

Our work complements that of \citet{Granule18} on the Granule programming
language.
Where Granule focuses on writing programs \emph{in} the language and running
them, we focus on metatheoretic reasoning \emph{about} type systems.
%As such, whereas Granule has a convenient syntax, performant type-checker,
%interpreter

Our work is similar in scope to that of \citet{LicataSR17}, though we work in
a natural deduction style rather than a sequent calculus style.
Where \citeauthor{LicataSR17} are much more agnostic in terms of
substructurality --- allowing for non-commutative and bunched logics ---
we are much more agnostic in terms of syntax.
The system of \citeauthor{LicataSR17} is essentially a single calculus,
supporting ``product'' ($\mathrm F$) types and ``function'' ($\mathrm U$)
types, parametrised on a \emph{mode theory} describing its structural rules.
For this system, they derive the strong result of cut elimination.
Meanwhile, we leave syntax design to the user, and consequently can only
guarantee substitution (which we can only get because of our commitment to
natural deduction).

@@

I carry out mechanised constructions and proofs in the proof assistant and
programming language Agda~\citep{Agda}.
Agda is based on Martin-L\"{o}f's intensional dependent type theory, so I
similarly present non-mechanised constructions and proofs assuming a foundation
given by dependent type theory, in a style inspired by the HoTT
Book~\citep{hottbook}.

The remaining three main chapters,
\cref{sec:framework,sec:semantics,sec:example-semantics}, adapt the syntactic
and semantic framework of \citet{AACMM21}, as presented at the end of
\cref{sec:simple}, to semiring-annotated calculi.
\Cref{sec:framework,sec:semantics} generalise the work on $\name$ presented in
\cref{sec:semirings,sec:ren-sub-lr}, respectively.
\Cref{sec:framework} shows how to formally describe the syntax of an arbitrary
semiring-annotated calculus, following the constructions used in
\cref{sec:semirings}.
\Cref{sec:semantics} then provides the generic environment-based semantic
traversal on such syntaxes, providing renaming and substitution as per
\cref{sec:ren-sub-lr} for all syntaxes as special cases of the generic
traversal.
\Cref{sec:examples} then gives further example uses of the generic traversal.

>>>>

This thesis advances the frontier of programming languages work that can be done
in a proof assistant.
I will elaborate on both of these aspects in the following paragraphs.

The main aim of programming language research is to find common patterns in
computer programs and represent these patterns via programming language
features.
As a basic example, consider functions in, say, a low level language like C, in
contrast to what these functions are compiled to in assembly language/machine
code.
In order to support function calls, the C runtime system manages a
\emph{call stack}.
A function call is compiled to a sequence of instructions which store the return
address and the values of the arguments on the stack, move the stack pointer,
and jump to the compiled code of the function.
A return from a function then places the return value into the appropriate
register, moves the stack pointer back, and jumps back to the return address.
Advances in programming language techniques are evaluated by the usefulness of
behaviours captured and goodness of the mathematical properties of the
abstractions produced.

Within programming language research, type theory is a methodology for producing
abstractions via classifying program behaviour.
The simplest and most common type systems classify programs by what values they
may return.
For example, the body of a C function with return type \texttt{char} is a
program that, if it returns a value, will return a character.
If that function has a parameter of type \texttt{int}, then we can compose it
with a function with return type \texttt{int} to build up a larger program.
We expect type systems to stop us from running programs with arguments of the
wrong types.
This holds of both static and dynamic type systems --- with a static type
system, the compiler will refuse to compile our code if we pass a \texttt{char}
value to a function expecting an \texttt{int}, while with a dynamic type system,
our program will do a check before running the function to make sure that we
have passed it an \texttt{int}.
The abstraction produced in such simple type systems is the idea that a function
will take arguments in accordance with its parameter types and produce a result
in accordance with its result type, and furthermore, as readers, we do not have
to inspect the function's implementation to see that these properties are true.

A more recent trend in type theory is to use types to describe other parts of a
program's behaviour than what range of values it may return.
For example, Java's \emph{checked exceptions} can be seen as part of the static
type system, in which programs are classified by which exceptions they may
throw.
More generally, \emph{effect systems} classify programs based on all of the
effects they may have --- such as reading input and writing to files, and also
internally defined effects, such as non-determinism and using an accumulator.
In a type system which tracks effects (an \emph{effect system}), programs by
default are pure (i.e.\ have no effects), with effects being opt-in.
Pure programs generally enjoy good properties, making them easy to reason about
and easy for an optimising compiler to optimise.

In this thesis, I consider the dual of effect systems ---
\emph{coeffect systems} --- as introduced by \citet{POM14}.
In a coeffect system, we are interested not in what extra behaviour a program
may exhibit (as with effects), but rather what extra abilities the context of a
program may provide.
Analogously to the case of effect systems, we typically restrict our
coeffect-free programs to be ``more pure'' than usual.
A standard example is to restrict to \emph{linear} programs, in which each
variable in the context is used exactly once.
The ability to duplicate and discard variables is then seen as a coeffect, which
can be tracked by a coeffect system.
Restricting to linear programs may seem like an arbitrary restriction at first,
but the expectation of linearity arises naturally in applications such as
file-handling, session-typed communication, and approaches to mutable memory.
I introduce linearity and its applications more fully in \cref{sec:linearity}.

%A type system lets us enforce in our language complex invariants which maintain
%abstraction boundaries while also ideally being machine-checkable.
%For example, we give functions function type to abstract away their
%implementation via call stacks and jumps and so on.

%In this thesis, the invariants of interest revolve around restricting the usage
%of variables.
%I introduce this topic thoroughly in \cref{sec:linearity}, but in brief\ldots

The work of this thesis relies upon type theory in two distinct ways.
Firstly, as I have introduced above, the main objects of study in this thesis
are programming languages with interesting type systems.
Secondly, however, type theory provides the basis of the proof assistant Agda I
use to implement the aforementioned programming languages and operations upon
them.
I will now introduce the idea of proof assistants.

A \emph{proof assistant}, also known as an \emph{interactive theorem prover}, is
a piece of software that allows for the encoding of mathematical definitions,
theorems, constructions, and proofs, and furthermore check that such encoded
proofs are correct and that such encoded constructions are well formed.
To truly by interactive, i.e.\ to actually assist, a proof assistant will
usually have a user interface which can read partial proofs, display
information about what more proof needs to be given, and provide actions that
will help complete the proof.

Proof assistants have seen increasing use in programming language research in
recent years.
The most obvious reason why working in a proof assistant is seen as beneficial
is that it ensures correctness.
If the proof assistant accurately implements a suitable mathematical foundation,
then any theorem proved in the proof assistant is guaranteed to be a true
theorem of that foundation.
These guarantees of correctness are particularly important when working with
combinatorially complex mathematical objects, proofs about which often require
the consideration of a large number of cases.
Programming language syntaxes are often such complex objects, motivating the use
of proof assistants when studying programming languages.

A second reason to use proof assistants is for the assistance they provide when
exploring a mathematical theory.
When we make a new definition, we may want to test how it works in a special
case, or what constructions it allows us to perform.
In a proof assistant, the assistance tools give us immediate feedback as to what
moves are and aren't allowed.
For example, if we define a complex type system, a proof assistant will let us
interactively build typing derivations, making clear any side conditions and
types of subderivations as we go.
Also, as I do later in this thesis, a proof assistant allows us to build a very
general theory, and practically use that theory directly in more specific cases
without losing rigour.

Thirdly, analogously to how a strong static type system can give us more
confidence when refactoring a program, the constant checking of proofs in a
proof assistant gives us the confidence to change definitions and lemmas knowing
that we will be guided towards the parts of our theory that need to be
correspondingly changed.
This can help if we are developing a new programming language with a changing
specification.

Finally, many proof assistants --- including Agda~\citep{Agda}, which I use in
this thesis --- double as programming languages themselves.
This means that we can write progams and prove properties of them using the same
tool.
Also, many theorems proven in such a proof assistant have computational content.
For example, if we prove a normalisation theorem for a programming language,
this will typically yield a (verified) normalisation algorithm for it, which we
can really run on a computer.
As such, a development in a proof assistant can provide a reference
implementation of a programming language, or even --- as with Idris
2~\citep{Brady21}, Lean 4~\citep{deMU21}, and Cedille~\citep{GRS16} --- the
actual implementation of a programming language.

The remaining three main chapters,
\cref{sec:framework,sec:semantics,sec:example-semantics}, adapt the syntactic
and semantic framework of \citet{AACMM21}, as presented at the end of
\cref{sec:simple}, to semiring-annotated calculi.
\Cref{sec:framework,sec:semantics} generalise the work on $\name$ presented in
\cref{sec:semirings,sec:ren-sub-lr}, respectively.
\Cref{sec:framework} shows how to formally describe the syntax of an arbitrary
semiring-annotated calculus, following the constructions used in
\cref{sec:semirings}.
\Cref{sec:semantics} then provides the generic environment-based semantic
traversal on such syntaxes, providing renaming and substitution as per
\cref{sec:ren-sub-lr} for all syntaxes as special cases of the generic
traversal.
\Cref{sec:example-semantics} then gives further example uses of the generic
traversal.

@@

I carry out mechanised constructions and proofs in the proof assistant and
programming language Agda~\citep{Agda}.
Agda is based on Martin-L\"{o}f's intensional dependent type theory, so I
similarly present non-mechanised constructions and proofs assuming a foundation
given by dependent type theory, in a style inspired by the HoTT
Book~\citep{hottbook}.
I give a fuller introduction to Agda in \cref{sec:agda-primer}.

This thesis is written in \colour{}, but should be readable without.
Agda code has syntax highlighting, and various pieces of notation related to
usage annotations are coloured \gr{green} for emphasis.

### 2023-07-16 tex/lenv.tex
[ writing ] address all remaining todos

\begin{example}
  We can derive the following instances of environment composition.
  \begin{description}
    \item[Composition of renamings.]
      If $\U = \V = \W = {\sqni}$, then we can take
      $\mathrm{lift}\,\rho\,x \coloneqq \rho(x)$.
      This allows us to compose two renamings into a single renaming.
    \item[Post-composition with a renaming.]
      More generally, if $\V = {\sqni}$ and $\U = \W$, we can still use
      the action of $\rho$ as $\mathrm{lift}$.
      This means that we can do a composition
      $\Gamma \env\U \Delta \env\sqni \Theta$ to get an environment of type
      $\Gamma \env\U \Theta$.
    \item[Composition of substitutions.]
      If $\V = \W = {\vdash}$, then $\mathrm{lift}$ is given by a
      syntactic traversal (in the sense of \cref{sec:syntactic-kits} and
      \cref{sec:lrsub}).
      For example, if $\U = {\sqni}$, we need the action of renaming on terms
      to show that a renaming followed by a substitution composes to a
      substitution.
      If $\U = {\vdash}$, then the action of substitution on terms gives us that
      substitutions compose.
    \item[Composing semantics with substitution.]
      More generally, if $\V = {\vdash}$ and we have a semantics (in the sense of
      \cref{sec:gen-sem} and \cref{sec:framework}) from
      $\U$ to $\W$, then $\mathrm{lift}$ can be given by the semantic traversal
      of terms.
  \end{description}
\end{example}

>>>>

We can derive the following corollaries as instances of environment composition.

\begin{corollary}[Composition of renamings]\label{thm:ren-comp}
  Given renamings $\rho : \Gamma \env\sqni \Delta$ and
  $\sigma : \Delta \env\sqni \Theta$, we can form their composite
  $\rho; \sigma : \Gamma \env\sqni \Theta$.
\end{corollary}
\begin{proof}
  Take $\U = \V = \W = {\sqni}$ in \cref{thm:env-comp}.
  Then let $\mathrm{lift}\,\rho\,x \coloneqq \rho(x)$.
\end{proof}

\begin{corollary}[Post-composition with a renaming]\label{thm:ren-env-comp}
  Given an environment $\rho : \Gamma \env\U \Delta$ and a renaming
  $\sigma : \Delta \env\sqni \Theta$, we can form their composite
  $\rho; \sigma : \Gamma \env\U \Theta$.
\end{corollary}
\begin{proof}
  As in \cref{thm:ren-comp}.
\end{proof}

\begin{corollary}[Pointwise renaming of an environment]\label{thm:env-ren}
  If $\sdtstile{}\V$ respects renaming, then so does $\env\V$ (on the left).
\end{corollary}
\begin{proof}
  Suppose we have $\rho : \Gamma \env\sqni \Delta$ and
  $\sigma : \Delta \env\V \Theta$.
  We want to compose these via \cref{thm:env-comp} with $U = {\sqni}$ and
  $\V = \W$.
  The function $\mathrm{lift}$ is given exactly by the fact that $\V$ respects
  renaming.
\end{proof}

\begin{corollary}[Composition of substitutions]\label{thm:sub-comp}
  Given substitutions $\rho : \Gamma \env\vdash \Delta$ and
  $\sigma : \Delta \env\vdash \Theta$, we can form their composite
  $\rho; \sigma : \Gamma \env\vdash \Theta$.
\end{corollary}
\begin{proof}
  Take $\U = \V = \W = {\vdash}$ in \cref{thm:env-comp}.
  Then, $\mathrm{lift}$ is given by the action of a substitution on a term
  (see \AgdaFunction{sub} in the following section).
\end{proof}

\begin{corollary}[Composing semantics with substitution]
  If we have a semantics (in the sense of \cref{sec:gen-sem} and
  \cref{sec:traversal}) from $\U$ to $\W$, then from an environment
  $\rho : \Gamma \env\U \Delta$ and a substitution
  $\sigma : \Delta \env\vdash \Theta$, we can form the composite
  $\rho; \sigma : \Gamma \env\W \Theta$.
\end{corollary}

### 2023-07-16 tex/linear.tex
[ writing ] various up to WRel example
[ writing ] address all remaining todos

The techniques of \cref{sec:simple} were developed to handle idealised core
calculi, like the simply typed $\lambda$-calculus and its many variants.
Such core calculi are fairly close to some real intermediate languages,
\todo{cite} though many intermediate languages also have polymorphism in their
type systems, which I have not discussed, but has been considered by much prior
work~\citep{POPLmark}.

Another potential feature --- required for many desired semantics but only
relatively recently appearing in programming languages --- that is not handled
by the techniques of \cref{sec:simple} is restrictions on how variables can be
used.
In the simple type systems covered in \cref{sec:simple}, the only restriction on
the use of a variable is that it is used as an expression of the appropriate
type.
However, we could also consider having somehow ``guarded'' subexpressions in
which some variables bound outside cannot occur, or systems in which the use of
a variable in one part of an expression precludes its use in another part of
that expression (the variable having been ``used up'').
I will introduce and motivate some such systems in this chapter, and argue that
they are impossible to capture directly in the framework presented in
\cref{sec:gen-syn}.

Languages and systems implementing features inspired by linear logic include
Rust~\citep{MK14,Rust}, ATS~\citep{Xi04,ZX05}, and various implementations of
session types~\citep{HLVCCDMPRTVTZ16}.
Additionally, forms of linearity have been used in theoretical work, for example
to bound computational complexity~\citep{GSS92,Hofmann03} or to write programs
which support incremental updates to computations~\citep{ER03,Ehrhard18}.
As for modality, \citet{IMO20} show that common data-flow analyses for
imperative languages can be recast as modal type systems, though they have
generally been developed in an ad hoc fashion.
Such a recasting may be useful to understand the metatheoretic properties of
programs passing such static analyses.
More user-facing implementations of modal type systems include the system for
stack allocation in OCaml~\citep{DW22} and the approach to programmer-annotated
erasure found in Agda and Idris~\citep{Atkey18}.

In this chapter, I look at two standard type-theoretic features that motivate
considering calculi going beyond simple type theories as deliniated in
\cref{sec:simple} or equivalently by \citet{AACMM21}.
The first feature is a $\Box$-modality --- specifically, the $\Box$-modality of
intuitionistic S4, which I discuss in \cref{sec:modal}.
The second feature is linearity, which I discuss in \cref{sec:ill}.
In particular, I discuss possible syntaxes for each, with a mind to
being able to apply the techniques described in \cref{sec:simple}.
Then, I\ldots

@@

\section{Mechanisations and systemitisations of substructural logics}
\label{sec:linmech}
\input{linmech.tex}

>>>>

The techniques of \cref{sec:simple} were developed to handle idealised core
calculi, like the simply typed $\lambda$-calculus and its many variants.
Such core calculi can be seen as simplified models of real programming languages
and/or their intermediate representations.
A potential programming language feature --- required for many desired semantics
but only relatively recently appearing in programming languages --- that is not
handled by the techniques of \cref{sec:simple} is restrictions on how variables
can be used.
In the simple type systems covered in \cref{sec:simple}, the only restriction on
the use of a variable is that it is used as an expression of the appropriate
type.
However, we could also consider having somehow ``guarded'' subexpressions in
which some variables bound outside cannot occur, or systems in which the use of
a variable in one part of an expression precludes its use in another part of
that expression (the variable having been ``used up'').
I will introduce and motivate some such systems in this chapter, and argue that
they are impossible to capture directly in the framework presented in
\cref{sec:gen-syn}.

Languages and systems implementing features inspired by linear logic include
Rust~\citep{MK14,Rust}, ATS~\citep{Xi04,ZX05}, and various implementations of
session types~\citep{HLVCCDMPRTVTZ16}.
Additionally, forms of linearity have been used in theoretical work, for example
to bound computational complexity~\citep{GSS92,Hofmann03} or to write programs
which support incremental updates to computations~\citep{ER03,Ehrhard18}.
As for modality, \citet{IMO20} show that common data-flow analyses for
imperative languages can be recast as modal type systems, though such analyses
have generally been developed in an ad hoc fashion.
Such a recasting may be useful to understand the metatheoretic properties of
programs passing such static analyses.
More user-facing implementations of modal type systems include the system for
stack allocation in OCaml~\citep{DW22} and the approach to programmer-annotated
erasure found in Agda and Idris~\citep{Atkey18}.

In this chapter, I look at two standard type-theoretic features that motivate
considering calculi going beyond simple type theories as deliniated in
\cref{sec:simple} or equivalently by \citet{AACMM21}.
The first feature is a $\Box$-modality --- specifically, the $\Box$-modality of
intuitionistic S4, which I discuss in \cref{sec:modal}.
The second feature is linearity, which I discuss in \cref{sec:ill}.
In particular, I discuss possible syntaxes for each, with a mind to
being able to apply the techniques described in \cref{sec:simple}.
After having introduced these topics, I finish this chapter with a survey of
prior work on representing such type theories in proof assistants in
\cref{sec:linmech}.

@@

\section{Mechanisations and systematisations of substructural logics}
\label{sec:linmech}
\input{linmech.tex}

### 2023-07-16 tex/linmech.tex
[ writing ] address all remaining todos

Also, the unusual form of sequents can cause some problems when working with a
typing with leftovers system.
For example, a traditional intuitionistic linear logic sequent
$\Gamma \vdash M : A$ corresponds to many different typing with leftovers
sequents:
\begin{itemize}
  \item $\gr1\Gamma \vdash M : A \boxtimes \gr0\Gamma$
  \item $\gr1\Gamma, \gr1x : B \vdash M : A \boxtimes \gr0\Gamma, \gr1x : B$
  \item $\gr1\Gamma, \gr0x : B \vdash M : A \boxtimes \gr0\Gamma, \gr0x : B$
  \item \&c.
\end{itemize}

Generally, any variable not used in the term can be added to both the input and
the output context with the same annotation.
Many of these variations are likely to appear in various typing derivations,
depending on what terms surround a given subterm.
This means that if we want to implement substitution, which involves putting a
term into an unknown surrounding, we have to navigate these different forms via
the \emph{framing property}.

@@

Such a scheme straightforwardly adapts to multiplicative linear logic by
modifying the data structures presented by \citet{McBride18} to disallow all
contraction and weakening.
With the additive rules, however, we get cases where variables appear multiple
times syntactically in a term but are still considered linear by the type
system, the simplest example being
$x : A \vdash \langle x, x\rangle : A \with A$.
Such rules are perhaps a new kind compared to what \citeauthor{McBride18}
considers, but it seems likely that just copying the same context to all the
premises would not be too hard to accommodate.
Meanwhile, we may consider implementing the $\oc$-modality as in DILL, with
intuitionistic variables handled using the regular co-de-Bruijn machinery from
the paper.
In summary, the co-de-Bruijn approach is promising for capturing linearity, but
has not been thoroughly investigated.

\subsection{Fitch-style modalities}

@@

\subsection{Systematisations of substructural logics}

@@

\paragraph{Granule}
\citet{BrunelGMZ14}, \citet{GhicaS14}, \ldots

>>>>

Also, the unusual form of sequents can cause some problems when working with a
typing with leftovers system.
For example, a traditional intuitionistic linear logic sequent
$\Gamma \vdash M : A$ corresponds to many different typing with leftovers
sequents, including:
\begin{itemize}
  \item $\gr1\Gamma \vdash M : A \boxtimes \gr0\Gamma$
  \item $\gr1\Gamma, \gr1x : B \vdash M : A \boxtimes \gr0\Gamma, \gr1x : B$
  \item $\gr1\Gamma, \gr0x : B \vdash M : A \boxtimes \gr0\Gamma, \gr0x : B$
\end{itemize}

Generally, any variable not used in the term can be added to both the input and
the output context with the same annotation.
Many of these variations are likely to appear in various typing derivations,
depending on what terms surround a given subterm.
In particular, many different variations can appear for the same term in
different subterms of the same derivation, even when the scope of the two
occurrences is the same.
This means that if we want to implement substitution, which involves putting a
term into an unknown surrounding, we have to navigate these different forms via
the \emph{framing property}.

@@

Such a scheme straightforwardly adapts to multiplicative linear logic by
modifying the data structures presented by \citet{McBride18} to disallow all
contraction and weakening, as presented by \citet{RPKV20}.
With the additive rules, however, we get cases where variables appear multiple
times syntactically in a term but are still considered linear by the type
system, the simplest example being
$x : A \vdash \langle x, x\rangle : A \with A$.
Such rules are perhaps a new kind compared to what \citeauthor{McBride18}
considers, but it seems likely that just copying the same context to all the
premises would not be too hard to accommodate.
Meanwhile, we may consider implementing the $\oc$-modality as in DILL, with
intuitionistic variables handled using the regular co-de-Bruijn machinery from
the paper.
In summary, the co-de-Bruijn approach is promising for capturing linearity, but
has not been thoroughly investigated.

The other relevant contribution from \citet{RPKV20} is to recast the
context-splitting relations via connectives inspired by bunched
logic~\citet{oHP99}.
I adapt this method in \cref{sec:lnd}, but with a different notion of
context-splitting.

\subsection{Fitch-style modalities}\label{sec:fitch}

@@

\subsection{Systematisations of substructural logics}\label{sec:lin-sys}

@@

\paragraph{Semiring-annotated systems}
There have been several appearances in the literature of calculi in which
variables are annotated with some algebraic usage information which is reflected
in the types of the calculus.
Such calculi appear in the work of \citet{abadi99core} and
\citet{reed10distance}, where annotations are used to control information flow
and sensitivity to perturbations, respectively.
Following these disparate calculi, the work of \citet{BrunelGMZ14},
\citet{GhicaS14}, and \citet{POM14} sought to unify these calculi using
(variations on) partially ordered semirings as the source of annotations.
Each of these papers also provided further examples of calculi with semiring
usage annotations.
I discuss calculi based on partially ordered semiring annotations further in the
following chapter, as it provides the basis for the rest of this thesis.

### 2023-07-16 tex/lr.tex
[ writing ] address all remaining todos

In this section, I introduce the syntax of the type theory \name{}, which makes
use of posemiring usage annotations to express the usage restrictions found in
DILL and other calculi.
I use this syntax to write some example programs, which will motivate the
denotational semantics explored in \cref{sec:wrel}.
For the rest of this thesis, particularly
\cref{sec:semirings,sec:ren-sub-lr,sec:wrel},
\name{} will serve as both a prototypical
usage-constrained syntax and a target of semantic analyses.

@@

I will not cover any operational semantics or equational theory of \name{} in
this thesis.
I will discuss a denotational semantics in \cref{sec:wrel}.

>>>>

In this section, I introduce the syntax of the type theory \name{}, which makes
use of posemiring usage annotations to express the usage restrictions found in
DILL and other calculi.
For the rest of this thesis, particularly
\cref{sec:semirings,sec:ren-sub-lr},
\name{} will serve as both a prototypical
usage-constrained syntax and a target of semantic analyses.

@@

I will not cover any operational semantics or equational theory of \name{} in
this thesis.
I will discuss a denotational semantics inspired by that of
\citet{AbelBernardy2020} in \cref{sec:den-sem}.

### 2023-07-16 tex/mech-related.tex
[ writing ] address all remaining todos

\citet{Autosubst18} propose some modifications to Autosubst which, as far as I
can tell, have not yet been incorporated, but are presented in mechanised form
for the paper.
Some of these modifications aim to bring Autosubst into line with
\citeauthor{AACMM21}'s work, in particular taking semantic traversals as a
theoretical basis.
TODO: more.

@@

Though \citet{AACMM21} did not state their results in categorical terms, it is
still useful to infer what the category-theoretic statement would have been, and
compare it to the statement actually given by \citet{FPT99} and \citet{FS22}.
When we do this, we see that while \citeauthor{FPT99} deal with the category of
contexts under renaming, and presheaves on that category.
This means that every model must be shown to respect renaming before touching
the framework of \citeauthor{FS22}.
Meanwhile, \citeauthor{AACMM21} make use of the interplay between the discrete
category of contexts and the category of contexts under renaming.
For example, the models (\AgdaRecord{Semantics}) of \citet{AACMM21} require only
that the family of semantic values $\V$ be shown to respect renaming, while the
fact that the result family $\C$ respects renaming follows as a corollary of the
traversal function \AgdaFunction{trav}.
\Citeauthor{FS22} make no distinction between $\V$ and $\C$, essentially only
having $\C$, but requiring it to respect renaming before getting the morphism
to that model from the initial model (the syntax).
In particular, this interplay allows us to derive renaming for terms in the way
we saw in \cref{sec:gen-syn} --- by making $\V$ the family of variables, which
trivially respects renaming.
\todo{Every model is an environment model?}

@@

In terms of the underlying theory, both works extend multi-sorted universal
algebra\todo{cite} with variable-binding.
However, the two extensions are subtly different.
Universal algebra already has a notion of variable, which supports renaming and
substitution, and which allows a given term to be evaluated when the free
variables of that term are assigned semantic values.
\citeauthor{AACMM21} reuse this notion of variable, and allow binding of such
variables in terms.
On the other hand, \citeauthor{FS22} recast the existing variables as
\emph{metavariables}, and introduce a new notion of (bindable) variables
separately into the syntax.
The resulting metavariables can then stand for arbitrary \emph{open} terms, and
thus each one remembers its context and requires an explicit substitution
whenever it is used.

@@

The other approach --- implementing a nominal library within an existing proof
assistant --- is exemplified by the Nominal Isabelle library of \citet{Urban08}.
This work is based around having a countably infinite discrete type
\texttt{name} (which could be a type of natural numbers or strings, for example)
which is treated abstractly, and then defining liftings of permutations on
\texttt{name} to permutations on any other types involving \texttt{name}s.
Because these liftings are defined explicitly within Isabelle/HOL, the Nominal
Isabelle setup amounts to an explicit model of a nominal logic within
Isabelle/HOL.
This setup resembles what \citet{AACMM21} did within Agda, and similarly what I
do later in this thesis --- creating a library within an existing proof
assistant.
The main downside, compared to working internally to a nominal set theory, is
that the external constructions of Nominal Isabelle are more complicated, and
involve keeping track of more properties that do not come as part of the theory
the proof assistant understands.

@@

From a broader perspective, I claim that nominal techniques solve a subtly
different problem to that solved by the likes of Autosubst, the work of Fiore,
and the techniques discussed in detail in this chapter.
The names found in nominal techniques are more general than (names of)
variables in formal systems.
This means that nominal techniques have been applicable to problems outside
standard programming language theory\todo{Find applications}.
However, in certain ways, this is arguably not a good fit for type systems.
In a basic nominal formalisation of a syntax with binding, there is no real
distinction between the context of a term and the collection of that term's free
variables.
This, for example, leaves no natural place to put the types of variables, except
as either a partial function from the set of names to types, or from the
computed set of free variables to types.
The discrepancy becomes even more apparent in substructural systems, as we see
later, where we want tight control over the context, as opposed to the scope.

>>>>

\citet{Autosubst18} propose some modifications to Autosubst which, as far as I
can tell, have not yet been incorporated, although all of the case studies in
the paper are mechanised in Coq.
In the paper, they adapt and extend the work of \citeauthor{ACMM17} on generic
semantic traversals to cover a variant of system F.
They use the term \emph{multivariate traversal} for the generalisation of
heterogeneous substitution to semantic traversals.
It appears that this work could be followed through to produce syntax
descriptions covering polymorphic calculi, which would provide a route for this
work to be incorporated into the Autosubst library.

@@

Though \citet{AACMM21} did not state their results in categorical terms, it is
still useful to infer what the category-theoretic statement would have been, and
compare it to the statement actually given by \citet{FPT99} and \citet{FS22}.
When we do this, we see that while \citeauthor{FPT99} deal with the category of
contexts under renaming, and presheaves on that category.
This means that every model must be shown to respect renaming before touching
the framework of \citeauthor{FS22}.
Meanwhile, \citeauthor{AACMM21} make use of the interplay between the discrete
category of contexts and the category of contexts under renaming.
For example, the models (\AgdaRecord{Semantics}) of \citet{AACMM21} require only
that the family of semantic values $\V$ be shown to respect renaming, while the
fact that the result family $\C$ respects renaming follows as a corollary of the
traversal function \AgdaFunction{trav}.
\Citeauthor{FS22} make no distinction between $\V$ and $\C$, essentially only
having $\C$, but requiring it to respect renaming before getting the morphism
to that model from the initial model (the syntax).
In particular, this interplay allows us to derive renaming for terms in the way
we saw in \cref{sec:gen-syn} --- by making $\V$ the family of variables, which
trivially respects renaming.
%\todo{Every model is an environment model?}

@@

In terms of the underlying theory, both works extend multi-sorted universal
algebra%\todo{cite}
with variable-binding.
However, the two extensions are subtly different.
Universal algebra already has a notion of variable, which supports renaming and
substitution, and which allows a given term to be evaluated when the free
variables of that term are assigned semantic values.
\citeauthor{AACMM21} reuse this notion of variable, and allow binding of such
variables in terms.
On the other hand, \citeauthor{FS22} recast the existing variables as
\emph{metavariables}, and introduce a new notion of (bindable) variables
separately into the syntax.
The resulting metavariables can then stand for arbitrary \emph{open} terms, and
thus each one remembers its context and requires an explicit substitution
whenever it is used.

@@

The other approach --- implementing a nominal library within an existing proof
assistant --- is exemplified by the Nominal Isabelle library of \citet{Urban08}.
This work is based around having a countably infinite discrete type
\texttt{name} (which could be a type of natural numbers or strings, for example)
which is treated abstractly, and then defining liftings of permutations on
\texttt{name} to permutations on any other types involving \texttt{name}s.
Because these liftings are defined explicitly within Isabelle/HOL, the Nominal
Isabelle setup amounts to an explicit model of a nominal logic within
Isabelle/HOL\@.
This setup resembles what \citet{AACMM21} did within Agda, and similarly what I
do later in this thesis --- creating a library within an existing proof
assistant.
The main downside, compared to working internally to a nominal set theory, is
that the external constructions of Nominal Isabelle are more complicated, and
involve keeping track of more properties that do not come as part of the theory
the proof assistant understands.

@@

From a broader perspective, I claim that nominal techniques solve a subtly
different problem to that solved by the likes of Autosubst, the work of Fiore,
and the techniques discussed in detail in this chapter.
The names found in nominal techniques are more general than (names of)
variables in formal systems.
This means that nominal techniques have been applicable to problems outside
standard programming language theory, such as in the representation of graphs
and in topology~\citep{GG17}.
However, in certain ways, this is arguably not a good fit for type systems.
In a basic nominal formalisation of a syntax with binding, there is no real
distinction between the context of a term and the collection of that term's free
variables.
This, for example, leaves no natural place to put the types of variables, except
as either a partial function from the set of names to types, or from the
computed set of free variables to types.
The discrepancy becomes even more apparent in substructural systems, as we see
later, where we want tight control over the context, as opposed to the scope.
